---
editor_options: 
  chunk_output_type: console
---

# Column names & classes 

## Setup

```{r}
#| warning: false
#| message: false
library(galah)
galah_config(atlas = "Australia") # default
galah_config(email = Sys.getenv("ALA_EMAIL"))

result <- galah_call() |>
  galah_identify("Litoria") |>
  galah_filter(year >= 2020, cl22 == "Tasmania") |>
  atlas_occurrences()
result |> head()
```


## Cleaning column names

* Info about using {janitor} and `clean_names()`?
* Info about renaming columns
* Info about splitting and/or joining columns?

## Column classes

We can check for obvious inconsistencies using the classes of each column. We
can do this with summary tables like a `skimr` report, or with base R. Below is
a simple example where our `decimalLatitude` column is numeric, which is what we
expect so in this case there is no problem. But as an example, if we change just
one of the values to a degrees minutes seconds format, we can see that the class
for the column changes to character. 

```{r}
sapply(result, class)
# Change one of the values to a degrees minutes seconds format
result$decimalLatitude[5] <- "40° 51' 59 N"
sapply(result, class)
```