---
editor: 
  markdown: 
    wrap: 72
---

# Missing values

Biodiversity datasets often contain missing values. However, analysis and visualisation software can differ in their default response to missing values. They might drop missing values, provide a warning about them, or return an error. We don't always know which response we'll get, which can cause problems for us (possibly without our knowing) during analysis later on. This is why it's important to clean missing values effectively. 

Understanding the nature of missing data can also be useful for informing how you proceed with analysis. For instance, understanding whether missing data are random or not can influence what you choose to do with them (e.g. remove vs. impute), which in turn can affect the results of statistical models based on those datasets. Identifying where missing data occur in a dataset is therefore an important part of exploratory data analysis. This chapter will show you how to detect and remove missing values.     

### Prerequisites

In this chapter, we will use gecko occurrence data since 2009 from the ALA using [galah](https://galah.ala.org.au/). 

```{r prereq}
#| message: false
#| warning: false
#| echo: false
library(galah)
library(dplyr)
library(tidyr)
galah_config(email = Sys.getenv("ALA_EMAIL"),
             verbose = FALSE)
geckos <- galah_call() |>
  identify("Gekkonidae") |>
  filter(year >= 2009) |>
  select(group = "basic",
         kingdom, phylum, order, class, 
         family, genus, species, cl22, 
         eventDate, month) |>
  atlas_occurrences()
```

```{r}
#| eval: false
# packages
library(galah)
library(dplyr)
library(tidyr)

# data: Gecko records since 2009
galah_config(email = "your-email-here") # ALA Registered email

geckos <- galah_call() |>
  identify("Gekkonidae") |>
  filter(year >= 2009) |>
  select(group = "basic",
         kingdom, phylum, order, class, 
         family, genus, species, cl22, 
         eventDate, month) |>
  atlas_occurrences()
```

:::{.aside}

<img class = "rounded" src="https://ala-images.s3.ap-southeast-2.amazonaws.com/store/8/7/9/8/89ad2446-e89c-481c-97d7-375ffad18978/original"></img>

::: {.figure-caption}
[*Heteronotia binoei* camouflaged over rocks, sticks & leaves.<br>Photo by Reiner Richter CC-BY-NC 4.0
(Int)](https://biocache.ala.org.au/occurrences/b0085bb1-72b9-4a4d-8dbf-05174b85c973)
:::
:::

## Find missing values

There are numerous ways to identify missing values in your data. Here we illustrate a few ways with very different output types.   

:::{.panel-tabset .nav-pills}  

#### `summary()`

The `summary()` function (base r) provides summary statistics for each column in your table. The output includes the number of missing values in each column (`NA's`).  Here, we can tell that there are missing values in the columns: `decimalLatitude`, `decimalLongitude`, `eventDate`, and `month`.  


```{r}
#| warning: false
#| message: false
summary(geckos) 
```


#### `skim()`  

The skimr package provides a fast and simple way to identify columns with missing values and quantify the amount of missing data. The output shows that there are 7 columns with missing values, shown in `skim_variable`, along with information about the number of missing values in `n_missing` and the proportion of non-missing values in `complete_rate`.  

```{r}
#| warning: false
#| message: false
library(skimr)

geckos |>
  skim() |>
  filter(n_missing > 0) |> 
  as_tibble()
```


#### `viz_miss()`  

The [visdat package](https://docs.ropensci.org/visdat/) contains functions to visualise different aspects of dataframes. `vis_miss()` allows users to visualise the extent and location of missing data throughout a dataframe, with additional arguments for customising the visual output through clustering or sorting the missing data. The output provides a visual summary of which columns have missing data and relatively how many rows have missing values.  

```{r}
#| warning: false
#| message: false
library(visdat)

vis_miss(geckos)
```

:::


From here we can return all rows with a missing variable.

```{r}
geckos |> 
  filter(if_any(everything(), is.na)) |>
  select(scientificName, genus, 
         species, cl22, everything()) # reorder columns
```

Or we can specify the column to find rows with `NA` values one or more columns in particular (in this case column `cl22`).

```{r}
geckos |> 
  filter(if_any(cl22, is.na)) |> 
  select(scientificName, genus, 
         species, cl22, everything()) # reorder columns
```

## Remove `NA` values

### Missing coordinates

If you are intending to make a map or run spatial analyses, it's a good idea to exclude records with missing coordinates. Missing coordinate data often leads to one of the following: a function error, an undesirable map, or the (unintentional) exclusion of data points (many of which you might wish to include in other maps or analyses). 

It's good practice to tally and identify rows that have missing data before excluding them. We can use the same method as above to identify the records with missing coordinates.

```{r}
geckos |> 
  filter(if_any(c(decimalLongitude, decimalLatitude), is.na)) |>
  count()

geckos |> 
  filter(if_any(c(decimalLongitude, decimalLatitude), is.na)) |>
  select(scientificName, decimalLongitude, 
         decimalLatitude, everything()) # reorder columns
```

If we decide we are happy to exclude these records, we can use `drop_na()` to remove missing values from our dataset. 

```{r}
#| message: false
#| warning: false
geckos_filtered <- geckos |>
  drop_na(decimalLongitude, decimalLatitude)
```

```{r}
#| echo: false
geckos |>
  drop_na(decimalLongitude, decimalLatitude) |>
rmarkdown::paged_table()
```


:::{.aside}

<img class = "rounded" src="https://ala-images.s3.ap-southeast-2.amazonaws.com/store/2/0/b/0/44f6694a-024a-45cd-896a-b73538fc0b02/original"></img>

::: {.figure-caption}
[*Gehyra dubia* camouflaged over rocks. <br>Photo by Nick Talbot CC-BY-NC 4.0
(Int)](https://biocache.ala.org.au/occurrences/996b962a-442d-412e-adad-59400bad3f6f)
:::
:::


We can check that `drop_na()` worked correctly by comparing the number of records in our initial data and our filtered data, specifically that `geckos_filtered` is smaller by the same number of `NA` values we found above.

```{r}
nrow(geckos) - nrow(geckos_filtered)
```


## Taxonomic values

### Missing higher rank

Sometimes data can be missing information in columns with higher taxonomic rankings, even if observations have been taxonomically identified down to the species level. In this case, the goal isn't only to remove the missing values, but to fill missing information with the correct information.

As an example, we'll use a slightly modified `geckos_missing` dataset with some missing data add into the higher columns[^1]. Below is a preview, and you'll notice that there are `NA` values in the `class` and `family` columns as you skim across the pages. 

[^1]: The Atlas of Living Australia matches taxonomic names when data is ingested from data providers. This process means that it's very rare for a species to be missing higher taxonomic rank names. This might not be the case for other data sources, though!


```{r geckos-missing-taxonomic-names}
#| warning: false
#| message: false
#| code-fold: true
set.seed(87654) # for reproducibility

# randomly replace some row's class & family names with NAs
geckos_missing <- geckos |>
  mutate(class = replace(class, 
                         sample(row_number(), 
                                size = ceiling(0.15 * n())), 
                         NA),
         family = replace(family, 
                          sample(row_number(), 
                                 size = ceiling(0.15 * n())), 
                          NA))
```


```{r}
#| echo: true
#| output: false
geckos_missing |>
  select(scientificName, class, family, genus, species, everything())
```


If you have a list of taxonomic names with complete ranking information from your preferred [taxonomic naming authority](../4_accessing/where-to-get-data.qmd#taxonomic-data), you can use this information to back-fill your missing data columns. In our case, we can use names from ALA's taxonomic backbone.

```{r}
#| echo: false
#| output: true
geckos_missing |>
  select(scientificName, class, family, genus, species, everything()) |>
  rmarkdown::paged_table()
```

First we'll get the complete list of *Gekkonidae* species taxonomic rank names using [galah](https://galah.ala.org.au/).

```{r}
#| echo: true
#| output: false
geckos_species_list <- galah_call() |>
  identify("gekkonidae") |>
  atlas_species()
```

```{r}
#| echo: false
#| output: true
galah_call() |>
  identify("gekkonidae") |>
  atlas_species() |>
rmarkdown::paged_table()
```


To match the complete names in `geckos_species_list` with the missing names in `geckos_missing`, we can use a strategy of match-and-merge. 

Using a reference column, we'll merge our complete names data to our missing data. The information in the `species_name` column of `geckos_species_list` contains both the genus and species names; this is just like the `scientificName` column of our `geckos_missing` dataset[^2]. We can use these columns as our reference columns.

<!--i'm a little confused as to why we are slicing the species_list and the _missing tibbles? is it just because you want to show an example with a small amount of data so it is digestible and less overwhelming? OT --> 

```{r}
geckos_species_list |> select(species_name) |> slice(10:15)
geckos_missing |> select(scientificName) |> slice(10:15)
```

Now we'll select the subset of columns from `geckos_species_list` that we wish to join with `geckos_missing` (and our reference column, `species_name`).

```{r}
list_subset <- geckos_species_list |>
  select(species_name, class, family)
```

We can use `left_join()` to merge `list_subset` to `geckos_missing`. Records that are identified at least down to the family level now have the correct rank information in the `class_new` and `family_new` columns [^3].

[^3]: This hasn't worked for every row, however, because `scientificName` contains the name of the lowest taxonomic rank the occurrence is identified to. Names like `Christinus` and `Gehyra` are genus names, so in these cases `scientificName` won't match our reference `species_name` column!

<!-- I think this footnote is kinda important to understand why there are NA's in the new columns. im thinking it would be better off not as a footnote especially because the article is about NA's explicitly it might raise some red flags at first glance of the table? :-) OT -->

```{r}
geckos_missing |>
  left_join(list_subset, 
            by = join_by(scientificName == species_name), 
            suffix = c("_old", "_new")) |>
  select(scientificName, class_old, family_old, class_new, family_new) |>
  slice(20:30)
```

<!--same here w the slicing OT --> 

### Insufficient taxonomic identification  

If a record is not identified down to the necessary taxonomic level required for your analysis (e.g. down to species or sub-species level), then the record should be removed.

One handy tip you can use with the [galah](https://galah.ala.org.au/) package is to add the column [`taxonRank`](https://dwc.tdwg.org/terms/#dwc:taxonRank). This column usefully shows the taxonomic level of names in `scientificName`[^2]. `taxonRank` can be useful because we can filter to only include records down to a specific rank.

[^2]: This column contains the name of the lowest taxonomic rank the occurrence is identified to.

```{r}
#| warning: false
#| message: false

geckos_rank <- galah_call() |>
  identify("gekkonidae") |>
  filter(year == 2013) |>
  select(taxonRank, scientificName) |>
  atlas_occurrences()

geckos_rank |>
  filter(taxonRank == "species")
```

However, it is still possible to filter records using the tools above without using the `taxonRank` column. In this case we remove records not identified down to the genus. 


```{r}
#| output: false
geckos_filtered <- geckos |>
  drop_na(genus) |>
  select(scientificName, genus, everything()) # reorder columns
geckos_filtered
```


:::{.aside}

<img class = "rounded" src="https://ala-images.s3.ap-southeast-2.amazonaws.com/store/c/d/a/d/817ae616-5596-407d-bab3-259ad78adadc/original"></img>

::: {.figure-caption}
[*Christinus marmoratus* camouflaged over lichen. <br> Photo by Edward CC-BY-NC 4.0
(Int)](https://biocache.ala.org.au/occurrences/69fda312-853a-4780-aeaf-9d6f7c7f47fe)
:::
:::
 

```{r}
#| echo: false
#| output: true
geckos_filtered |>
rmarkdown::paged_table()
```
 
## Summary

This chapter showed ways to find and remove different types of missing records from your dataset. Importantly, there might not be one single cleaned dataset missing all missing values. You might find that in the same workflow, you will clean your dataset in multiple ways to preserve rows that are missing some, but not all, information. For example, some records have complete taxonomic identification but lack spatial coordinates. Other records have spatial coordinates but lack taxonomic information. Depending on your intended type of analysis, you may have to vary how you clean your data for each analysis!

In the next chapter we will address working with **strings** (character sequences). Strings often involve correcting for things like typos or extra spacing that are difficult to pick up at a first glance.

