---
editor_options: 
  chunk_output_type: console
---

# General utility functions

<!-- Notes/Andrew: Just some mostly incomplete stuff I think is basic or else
doesnt have a home elsewhere yet. Some concepts are like what was in the old
precleaning page, but precleaning isn't a real thing its just cleaning -->

Here we will explore a few simple ways to identify and rectify the more obvious
or common sources of inconsistency.

## Setup

```{r}
library("galah")
galah_config(atlas = "Australia") # default
galah_config(email = Sys.getenv("ALA_EMAIL"))
```

```{r}
result <- galah_call() |>
  galah_identify("Litoria") |>
  galah_filter(year >= 2020, cl22 == "Tasmania") |>
  atlas_occurrences()
result |> head()
```

## Dates

Some use cases may require dates beyond a simple year value. Standardising dates
involves ensuring that the variables in a dataset have values that conform to a
consistent and standard format.  An example of unstandardised data is having
varied date formats (e.g. DD/MM/YYYY for some entries and MM/DD/YYYY for
others). This may be necessary when integrating data from multiple sources, but
it is important to remember that there can be inconsistencies even within a
single source dataset.

## Column classes 

We can check for obvious inconsistencies using the classes of each column. We
can do this with summary tables like a `skimr` report, or with base R. Below is
a simple example where our `decimalLatitude` column is numeric, which is what we
expect so in this case there is no problem. But as an example, if we change just
one of the values to a degrees minutes seconds format, we can see that the class
for the column changes to character. 

```{r}
sapply(result, class)
# Change one of the values to a degrees minutes seconds format
result$decimalLatitude[5] <- "40° 51' 59 N"
sapply(result, class)
```

## Unexpected values

* Checking for unexpected values: this is a generic method but the resolution
logic depends on the issue (taxonomic, categories, strings, etc.)
  * Context: a merged dataset pertaining to a single species (using data frame
  from cleaning_integration.qmd). Species is L. chloris
    * Assumption: species column contains only one species
      * Method: `unique(merged_data$species)`
      * Result: two species names
      * Resolution: conform to one species name (assign)
    * Assumption: country code contains only one country
      * Method: `unique(merged_data$country_code)`
      * Result: AU, NA, JP
      * Resolution: Investigate NA and assign, investigate JP since chloris is
      an Australian species

```{r}
library(dplyr)
merged_data <- read.csv("data/galah/chloris.csv")

unique(merged_data$countryCode)

merged_data[which(merged_data$countryCode == "JP"), ]

# where should the point be? can check the `locality` column and coordinates

merged_data[which(merged_data$countryCode == "JP"), ]$decimalLatitude
merged_data[which(merged_data$countryCode == "JP"), ]$decimalLongitude
merged_data[which(merged_data$countryCode == "JP"), ]$locality

# mt bucca is in australia but the coordinates are incorrect
# the latitude is missing an "-"
# we can fix this and check the result (#TODO map vis)
fixed <- merged_data %>%
  mutate(decimalLatitude = ifelse(countryCode == "JP", paste0(
    "-",
    decimalLatitude
  ), decimalLatitude)) %>%
  mutate(countryCode = ifelse(countryCode == "JP", "AU", countryCode))
```

## Summary

In this chapter, we learned a few basic checks for cleaning datasets, including
methods to detect inconsistencies in date formats, coordinate systems, and
units.

