---
editor: 
  markdown: 
    wrap: 72
number-depth: 3
code-annotations: hover
execute: 
  eval: false
---

# Data in this book {.unnumbered}

This book uses data DOIs to preserve reproducible data downloads, as most data is downloaded from the [Atlas of Living Australia (ALA)](https://ala.org.au/) using the [galah package](https://galah.ala.org.au/R/), the ALA's R interface to downloading biodiversity data. At the beginning of each chapter, in the *Prerequisites* section, we have provided options for you to either download the identical dataset using these DOIs or reproduce the query using our code.  

::: {.callout-note collapse="true" appearance="simple"}

#### Using original download queries

If you download data using a galah query (rather than a DOI), the results you return when running the examples might not exactly match results in this book. This difference arises because data in the ALA is constantly being ingested (including from previous years) and galah downloads data directly from the ALA. This means that using the same query after the date this book was last rendered---even a query for a specific time-period (e.g., 2010--2015)---*might* return different data today. 

In general, that is a good thing; updates improve data quantity and/or quality. But they can affect reproducibility, too. Getting the latest data shouldn't affect whether the example works correctly, but it may affect whether the results you return in your console match the results returned in this book. The main advantage to using *Prerequisites* sections is that data downloads are transparent and up-to-date.

:::

There are a few datasets used in this book that were not retrieved from the ALA. To ensure reproducibility, we have provided download links to those datasets in this chapter. 


### Prerequisites

```{r}
#| warning: false
#| message: false
library(arrow)
library(dplyr)
library(readr)
library(here)
```

## Data

### `bees`

This is a subset of a larger dataset---[Curated Plant and Invertebrate Dataset for Bushfire Modelling](https://data.csiro.au/collection/csiro:56679). 

::: aside

<img class = "rounded" src="https://ala-images.s3.ap-southeast-2.amazonaws.com/store/0/0/e/5/277a5c28-827a-49c7-8588-d16333925e00/original"></img>

::: {.figure-caption}
[*Braunsapis* species hovering in front of a ghost gum tree flower.<br>Photo by Zig Madycki CC-BY-NC-ND 4.0 (Int)](https://biocache.ala.org.au/occurrences/92928258-92a1-4e6b-b9e2-46da0d8df383)
:::
:::

Click on the download button below to download as a parquet file. Save the parquet in your working directory and load it using the code below.

{{< downloadthis data/dap/bees.parquet dname="bees" label="Download bees data" icon="database-fill-down" id=download-button >}}

```{r}
#| eval: false
#| warning: false
#| message: false
bees <- read_parquet(here("path", "to", "bees.parquet"))
```


### `inverts`

This is a subset of a larger dataset---[Curated Plant and Invertebrate Dataset for Bushfire Modelling](https://data.csiro.au/collection/csiro:56679). 

::: aside

<img src="https://ala-images.s3.ap-southeast-2.amazonaws.com/store/0/5/6/a/ca48dfdd-15b3-4ac5-a392-cf5f93a6a650/original" class="rounded"/></img>

::: figure-caption
[*Paralaoma mucoides* on a rock.<br>Photo by Nick Porch CC-BY-NC 4.0
(Int)](https://biocache.ala.org.au/occurrences/8a44a8df-2819-45d4-b8c5-e6dc7629e0fe)
:::
:::

Click on the download button below to download as a parquet file. Save the parquet in your working directory and load it using the code below.

{{< downloadthis data/dap/inverts_subset.parquet dname="inverts" label="Download inverts data" icon="database-fill-down" id=download-button >}}

```{r}
#| eval: false
#| warning: false
#| message: false
inverts <- read_parquet(here("path", "to", "inverts.parquet"))
```


### `gbif_species_list`

A species list of *Eucalyptus* downloaded from GBIF.

Click on the download button below to download as a parquet file. Save the parquet in your working directory and load it using the code below.

{{< downloadthis data/lists/gbif_eucalyptus.parquet dname="gbif_species_list" label="Download gbif_species_list data" icon="database-fill-down" id=download-button >}}

```{r}
#| eval: false
gbif_species_list <- read_parquet(here("path", "to", "gbif_species_list.parquet"))
```


### `marine_sp`

A subset of World Register of Marine Species (WoRMS) data.

Click on the download button to download as a .csv file. Save the csv in your working directory and load it using the code below.

{{< downloadthis data/worms/worms.csv dname="marine_sp" label="Download marine_sp data" icon="database-fill-down" id=download-button >}}

```{r}
#| eval: false
marine_sp <- read_csv(here("path","to", "worms.csv"))
```

