# Taxonomy {#sec-standardise-taxonomy}

New species are being discovered and described all the time. The rapid rise of analyitical tools, especially in molecular biology can make it difficult to keep up with taxonomic literature. Species that historically were described solely on morphological characteristics are being re-classified post genomic/population genetics, and integrative taxonomy, an approach which integrates information from multiple areas which can be used to delimit and describe new species [@garraffoni2019integrative].

Taxonomy is dynamic, data can therefore be inconsistent between databases and across timescales. Additionally the scientific community do not always agree on the classification of species, this results in multiple phylogenetic trees for the same species or arrays of subspecies.

## Taxonomy in biodiversity databases  

##note this section might be best placed somewhere else i'm not sure

To deal with naming inconsistencies, naming authorities are used by online biodiversity databases in order to classify species \[REF\]. Different databases might use different naming authorities, you might not agree with their classifications, and there may be other issues you are not aware of:

-   people assume the taxonomic backbone the ALA uses is correct

-   assume that data has been parsed

-   this process is not transparent

-   taxonomic backbone changes without documentation/record, if you re-download data for example a year later things may have changed without traceable decisions.

Errors less likely at higher level taxonomy than at the species level. th

so how do you deal with taxonomic inconsistencies?

## Naming authorities

### What is a naming authority

**I think it's worth actually defining this very clearly because its not common knowledge**

Using a naming authority is a great way to make decisions surrounding taxonomic categorizations of the data you're working with. Accurate species delimitation is crucial for adequate conservation management and understanding evolutionary processes \[REF\]. Species-level lists re the foundation of conservation decisions, such as is the IUCN Red List [@melville21]. However, deciding what naming authority to use can be challenging and time consuming.

### Choosing a naming authority

Choosing a naming authority will be dictated by whether or not you are looking at a single species or maybe a genus, or if you're looking at all species in a certain region. The difference in scope might influence if you choose to use a naming authority from a taxonomic society group or something that covers for example all plants.

A knowledge of taxonomy is essential in order to verify the data you are working with. For this, different organizations provide updated lists of species or even details of the taxonomic history of a species. Checking changes in taxonomy can be helpful when interpreting old data which may have species names you don't recognize. Changes in taxonomy, such as species split, new higher level classification (as genera), or species descriptions, highlight the importance of keeping up-to-date with taxonomic literature. This can be achieved by consulting the literature. Most taxonomic society groups release annual updates on taxonomy. The [Atlas of Living Australia (ALA)](www.ala.org.au) can also be used to investigate taxonomic changes.

In Australia, the [Australian Plant Name Index (APNI)](https://www.anbg.gov.au/apni/) is the primary naming authority for plants. With the [Australian Faunal Directory (AFD)](https://biodiversity.org.au/afd/home) the main taxonomic catalog for animal species. These authorities provide a list of accepted and authoritative names as a template. If you're unsure what naming authority to use, APNI and the AFD are a good place to start, especially if the data you're investigating covers a wide range of taxa. If you're investigating certain taxa it's worth checking when the taxonomy was last updated in the APNI or AFD, especially if you know there has been recent changes. If you want to investigate closer, we've provided some links to society groups, in some cases these can be more up to date that the APNI or AFD.

### Australian taxonomic society groups

**VERTEBRATES**

-   Amphibians and reptiles - [Australian Herpetological Society](https://ahs.org.au/)\
-   Birds - [Birdlife Australia](https://www.birdlife.org.au/)\
-   Fish - [Australian Society for Fish Biology](https://www.asfb.org.au/)\
-   Mammals - [The Australian Mammal Society](https://australianmammals.org.au/)

**INVERTEBRATES**

-   Arachnology - [Australasian Arachnological Society](www.australasianarachnologicalsociety.org)\
-   Entomology - [Australian Entomological Society](https://www.austentsoc.org.au/)\
-   Malacology - [The Malacological Society of Australasia](https://www.malsocaus.org/)\
-   Nematology - [Australasian Association of Nematologists](https://www.nematologists.org.au/)

## What to do with the naming authority

Now that you've chosen a naming authority you'll want to use it to ensure the data you're working with has consistent taxonomy

```{r}

```

## Synonymy

What are synonyms

Synonyms can be annoying to deal with when working with open source data, and they can appear because of multiple reasons

1.  Taxonomy changes over time, a record from the 1950s might have a name that is no longer used for the species
2.  The scientific community doesn't always agree, there may be multiple names, and taxonomic classifications for the same species

There are two main types of synonyms when it comes to taxonomy: **heterotypic** and **homotypic.** The type of synonym might dictate how you handle it.

Often naming authorities will have one they deem as accepted, and list others as synonyms.

Note: maybe need to explain difference between **heterotypic** and **homotypic** synonyms

## Standardize taxonomy

The following steps assume that the taxonomic history of the included species is known and correct and out-of-date names and synonymy have been fixed.

Ensure that the spelling is correct and fix errors as they appear. It is essential to decide on a standard way to record the scientific names and be consistent. For example, start with upper case and separate the names with an underline e.g. "Moloch_horridus".

It is also important to check the taxonomic hierarchy, make sure it's correct and nothing is missing, such as family, and order. You can often fill this information in using your chosen naming authority or referring back to an atlas such as the ALA.

```{r}

```

It's best practice to standardise and fix what you can before you start removing anything, this decreases the chance of removing more records than you need to.

-   Remove non-identifiable specimens: If a species is not discernible after checking spelling and taxonomy, then it should be removed.

-   Remove records with insufficient taxon identification: If a record is not identified down to the species level and this level of detail is needed for the study, then the record should be removed.

-   Remove records with miss-identified taxonomy: Incorrectly identified species, or unrecognized species names compared to a naming authority, should be removed. Incorrectly identified specimens can be difficult to identify with open source biodiversity data. Often these will be picked up by 1) an image of the species in question which does not match 2) If you notice a species outside of its geographic range, this could be a true outlier, it could be a spatial error, or it could be a different species. (**see --- for more info**)

```{r}

```

### Optional taxonomic cleaning

The following steps may not apply to your research interests, however this type of cleaning is commonly required.

-   Remove non-native species: This step is a common requirement. A list can be obtained from the [Global Register of Introduced and Invasive Species (GRIIS)](https://griis.org/)

-   Remove domesticated specimens, sightings of dogs, cats, but also garden species (**elaborate**)

-   Remove extinct species: if you don't have a year filter on your data you might have records of species that have become extinct in more recent years.

-   Remove specific taxa/life stages depending on the study: For example, if working with terrestrial data, it is necessary to remove marine taxa.

```{r}

```

## 
