# Precleaning 

## Familiarize with the data and metadata

Now that you have data to work with it's important to familiarise yourself with it. This is best done before you get started with any data cleaning [@streamdna_sharing_2020].

-   Metadata can provide information on the structure of the data and how the data was collected.

-   The metadata may also outline or provide insights into the limitations of the data.

-   A visual inspection of your entire dataset can save time and solve easy-to-spot errors.

-   Help identify limitations

A simple way to visualize your data is to plot it on a map. This simple step can highlight coordinates with missing data or even inverted or erroneous coordinates (as points floating in the ocean while focusing on terrestrial species and vice versa).

Other tools, such as looking for only positive values (if that is what is expected) can also be used in this initial data familiarization step.

```{r}

```


Systematic data cleaning is required, however, some mistakes observed during the initial visualization of the data and metadata can be fixed straight away.

-   Typos

-   Ensure consistent capitalizations

-   Fix minor coordinates errors, such as inverted or badly formatted

-   Remove records with no coordinates

```{r}

```