---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Temporal scope

A good first step to understanding what data are available is to check how many
observations there are over different time periods. We can check how many total
observations there are in the Atlas of Living Australia (the largest
biodiversity data aggregator in Australia) with the following query:

```{r} library(galah) ```



```{r} #| warning: false #| message: false library(galah)

atlas_counts() ```

Now let's check how many total insect records there are.

```{r} galah_call() |> galah_identify("insecta") |> atlas_counts() ```

Now let's see how many insects have been observed each year over the last 10
years. We'll order this by descending year using `dplyr::arrange()` and
`dplyr::desc()`.

```{r} #| warning: false #| message: false library(dplyr)

galah_call() |> galah_identify("insecta") |> galah_filter(year >= 2013) |>
  galah_group_by(year) |> atlas_counts() |> dplyr::arrange(dplyr::desc(year))
  ```

Now that we have an idea of the total amount of data in the Atlas of Living
Australia, let's try checking how many observations exist of the genus
*Lampromicra*. We'll first make sure the scientific name *Lampromicra* returns
the taxon information we expect with `search_taxa()`.

```{r} search_taxa("Lampromicra") ```

This looks correct! Next let's see how many total observations there are of
*Lampromicra* and how many observations there were in each of the last 10 years.

```{r} galah_call() |> galah_identify("Lampromicra") |> atlas_counts()

galah_call() |> galah_identify("Lampromicra") |> galah_filter(year >= 2013) |>
  galah_group_by(year) |> atlas_counts() |> dplyr::arrange(dplyr::desc(year))
  ```

There are a growing number of observations from 2012 to 2023 of Jewel beetles in
the Atlas of Living Australia, with fewer than 100 observations each year prior
to 2020. 

With this information, we might choose to combine data from all years in our
analysis (rather than separating them by year). Alternatively, we might decide
to only include data since 2020, which might be sufficient to represent where
*Lampromicra* are found and be more relevant because they are more recent
observations. These are decisions that might affect the granularity of the
research question we ask.

