
### Extended taxonomic cleaning

Finally, we will end with a worked example that shows the application of these approaches when consolidating data from different data providers.


(I think this should be in another section)

The following steps may not apply to your research interests, however this type of cleaning is commonly required.

-   Remove non-native species: This step is a common requirement. A list can be obtained from the [Global Register of Introduced and Invasive Species (GRIIS)](https://griis.org/)

-   Remove domesticated specimens, sightings of dogs, cats, but also garden species (**elaborate**)

-   Remove extinct species: if you don't have a year filter on your data you might have records of species that have become extinct in more recent years.

-   Remove specific taxa/life stages depending on the study: For example, if working with terrestrial data, it is necessary to remove marine taxa.

```{r}

```

## Merge datasets from different infrastructures

(Park this for now may come after naming authority)

If you have downloaded data from different sources, you likely will need to collate your data into a singular database.

When combining data from multiple places, it is important to standardize the data fields and merge the data carefully [@ribeiro_bdc_2022]. There is a chance data will be incorrectly formatted and/or mislabeled.

One of the main issues you might face if you've sourced data from different organisations/people is that higher taxonomy may not match- if this is the case, take a look at the "taxonomy" chapter for more information on how to deal with inconsistent taxonomy before merging!

```{r}
##example on formatting consistency, merging data together etc
```

