---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Taxonomic scope

Taxonomy refers to ways by which we classify an organism and its relationship to
all other organisms. Typically, your research question will be concerned with
one or more taxonomic groups, ranging from a single species to an entire kingdom
(e.g. *Plantae*). 

Following on with our example, we are interested in understanding more about
Jewel beetles in the genus *Lampromicra*.

We first might want to know what species there are in the genus *Lampromicra*
and view some additional taxonomic information about them. We can do this by
adding `atlas_species()` to the end of a query in {galah}.

:::{.callout-note} 

You will need to add an email address [registered with the
Atlas of Living
Australia](https://auth.ala.org.au/userdetails/registration/createAccount) in
`galah_config()` to download species information. 

:::

```{r} 
library(galah)
```

```{r}
#| eval: false
# Provide your registered ALA email address
galah_config(email = "youremail@here.com")
```

```{r}
#| include: false
galah_config(atlas = "Australia") # default
galah_config(email = Sys.getenv("ALA_EMAIL"))
```

```{r}
galah_call() |>
  galah_identify("Lampromicra") |>
  atlas_species() |>
  select(family:species_guid)
```

Our query returned three species in the genus *Lampromicra*. The URLs returned
under `species_guid` can be entered in a web browser to see more information. 

We might also wish to check the total number of observations of each of these
species. We can check this by grouping our counts by species using
`galah_group_by(species)`

```{r} 
galah_call() |>
  galah_identify("Lampromicra") |>
  galah_group_by(species) |>
  atlas_counts()
```

Our result shows that the majority of observations of *Lampromicra* are of the
species *Lampromicra senator*. 

Given this information, we might consider whether our question needs to be made
at the species level, or whether we might increase the taxonomic scope to the
genus or family level to use more data. Ultimately, the taxonomic scope of your
data will depend on how important it is to your question to compare specific
taxonomic groups.
