---
editor_options: 
  chunk_output_type: console
---

# Taxonomic validation

## Taxonomic classification

Taxonomic classification is a complex issue when working with open source
biodiversity data. Data infrastructures have their own taxonomic systems, which
can lead to variations across platforms in hierarchical information.
Furthermore, taxonomic classification is in a state of constant change, and
views may differ within the literature or across authorities (refer back to
Taxonomy and open source data). As a result, it is
often the case that you will have synonyms, meaning, more than one name for the
same species, in your dataset. Keep in mind that there is no universal solution
to these issues. However you choose to resolve them, we recommend that you
maintain a clear and explicit record of any decisions and changes made with
respect to the data.

### Detecting synonyms

There are several packages available that can be used to query different
taxonomic databases and check for synonyms. 

#### `worrms`  

The [{worrms}](https://docs.ropensci.org/worrms/) is the R interface to the
World Register of Marine Species and has the ability to cross check synonyms
with their database for taxa that has an `AphiaID`. The function will return
synonymous record(s) associated with another different AphiaID. 

```{r}
#| warning: false
#| message: false
library(worrms)
library(tidyverse)

marine_sp <- read_csv("../data/worms/worms.csv")

marine_sp |>
  slice(7) |>
  pull(AphiaID) |>
  wm_synonyms()
```

#### `taxize` 

<!-- Notes/Andrew: There's no explanation for anything about this list here,
what is it, where is it from etc -->
<!-- Notes/Andrew: Wouldn't it be better to show, or also show, how to do this query with your dataset? -->
The `taxize` package allows users to search over many taxonomic data sources for
hierarchial taxonomic information, such as species names (scientific and
common), to resolve synonymy. The `gnr_resolve()` function matches a supplied
list with up to 118 data sources including GBIF, Catalogue of Life, World
Register of Marine Species and many more. The function scores how well matched
your name is to these sources. 

```{r}
library(taxize)

# Read in a naming authority list
afd <- read_csv("../data/naming/afd.csv")
unique(afd$VALID_NAME)

# Resolve names
resolved <- gnr_resolve(unique(afd$VALID_NAME), best_match_only = TRUE)
resolved |> print(n = 50)

# Retrieve synonyms
tsn <- get_tsn(unique(afd$VALID_NAME)[1:5])
synonyms(tsn)


```

## Input from experts

Programmatic solutions for resolving synonymy can only go so far. Seeking
validation from experts is sensible if your goal is to obtain a high quality
species list. Museums or taxonomic societies are extensive sources of knowledge.
Below we have provided a list of some of Australian taxonomic society groups. 

### Australian taxonomic society groups

**VERTEBRATES**

-   Amphibians and reptiles - [Australian Herpetological Society](https://ahs.org.au/)\
-   Birds - [Birdlife Australia](https://www.birdlife.org.au/)\
-   Fish - [Australian Society for Fish Biology](https://www.asfb.org.au/)\
-   Mammals - [The Australian Mammal Society](https://australianmammals.org.au/)

**INVERTEBRATES**

-   Arachnology - [Australasian Arachnological Society](www.australasianarachnologicalsociety.org)\
-   Entomology - [Australian Entomological Society](https://www.austentsoc.org.au/)\
-   Malacology - [The Malacological Society of Australasia](https://www.malsocaus.org/)\
-   Nematology - [Australasian Association of Nematologists](https://www.nematologists.org.au/)

### Global taxonomy

-   [GBIF taxonomic backbone](https://www.gbif.org/dataset/d7dddbf4-2cf0-4f39-9b2a-bb099caae36c) - Uses over 100 different sources
-   [Integrated Taxonomic Information System, ITIS](https://www.itis.gov/) - Authoritative taxonomic information on plants, animals, fungi, and microbes
-   [Catalogue of Life](https://www.catalogueoflife.org/) - Global taxonomic catalogue   
