# Large downloads

Once the size of your dataset starts to approach the size of your computer's memory (or vastly exceeds it!), it becomes difficult to process this data using R. There are a few different approaches to managing this, such as using the [{arrow}](https://arrow.apache.org/docs/r/index.html) package for larger than memory data workflows or the [{duckdb}](https://r.duckdb.org/index.html) package for querying databases using SQL.  

### {arrow}  
The {arrow} package allows users to easily read and write different file formats, and analyse larger than memory datasets using {dplyr} verbs. A common workflow consists of opening a dataset, filtering or summarising according to some property of the data, and writing this to a new file. This is relatively simple to do with a small csv file, and we can do something conceptually similar using {arrow} for larger datasets. 

```{r}
#| message: false
#| warning: false
library(arrow)
library(dplyr)
library(readr)
```

Let's start by getting a large dataset: this 2GB biodiversity dataset from the [EcoAssets project](https://ecoassets.org.au) contains 11,619,897 rows of counts of species from 1900-2022. More details are available [here](https://ecoassets.org.au/data/aggregated-data-australian-species-occurrences/). 

```{r}
# figure out a way to direct users to download this dataset somewhere
```


```{r}
# how to get small dataset 
# just documenting this so I don't forget
# species_counts |> 
#   filter(year %in% c(2000:2001)) |>
#   write_csv_arrow("data/ecoassets/species_counts_small.csv")
```

**Maybe tabs to show dplyr/arrow workflow side by side?**

```{r}
# open dataset
species_counts <- open_dataset(here::here("data", "ecoassets", "species_counts.csv"), format = "csv")

# vs
species_counts_small <- read_csv(here::here("data", "ecoassets", "species_counts_small.csv"))
```

Summarising a large dataset using the {arrow} package consists of all the same steps you would follow for a smaller dataset using {dplyr} functions (e.g. selecting, filtering, grouping by, summarising), with the addition of a line at the end of the workflow to `collect()` the output back into memory. 

```{r}
# filter, group_by, summarise, collect
epbc_summary_2000 <- species_counts |> 
  select(year, epbcStatus, speciesName) |> 
  filter(year == 2000) |> 
  group_by(epbcStatus) |> 
  summarise(unique_species = n()) |> 
  collect()
  
# vs
epbc_summary_2000_small <- species_counts_small |> 
  select(year, epbcStatus, speciesName) |> 
  filter(year == 2000) |>
  group_by(epbcStatus) |> 
  summarise(unique_species = n()) 

```

If the 

This is very convenient when the operations we want to perform have perfect mappings between {dplyr} and {arrow}, but what if we want to transform the large dataset in a way that is not (currently) supported by {arrow}?

```{r}
species_counts |> 
  select(year, epbcStatus, speciesName) |> 
  filter(year == 2000) |> 
  group_by(epbcStatus) |> 
  summarise(unique_species = n()) |> 
  collect()
```


```{r}
# convert to parquet files and wrangle
```

**should we link to the arrow help pages for each of these sections? so they don't seem outdated as things progress? discuss with Dax**

### {duckdb}
