---
editor: 
  markdown: 
    wrap: 72
number-depth: 3
code-annotations: hover
---

# Geospatial investigation {#sec-geospatial-cleaning}

An important part of observational data is the location, specifying where each observation of an organism or species took place. These locations can range from locality descriptions (e.g. “Near Tellera Hill station”) to exact longitude and latitude coordinates tracked by a GPS system. The accuracy of these geospatial data will determine the types of ecological analyses you can perform. It is important to know the precision of these observations, along with the range of uncertainty around an observation’s location, to contextualize any findings or conclusions made using the data.

In this chapter, we will discuss some different ways to assess the precision and uncertainty of coordinates associated with occurrence records, and highlight how to identify spatial characteristics or errors when visualizing occurrences on maps.

### Prerequisites

In this chapter we'll use data of *Banksia serrata* occurrence records since
2022 and quokka occurrence records from the ALA.

```{r prereq-internal}
#| echo: false
#| warning: false
#| message: false
# packages
library(galah)
library(ggplot2)
library(dplyr)
library(sf)
library(ozmaps)
library(tidyr)
library(stringr)
galah_config(email = Sys.getenv("ALA_EMAIL"),
             verbose = FALSE)

# get DOIs for chapter
doi_table <- here::here("data", "galah-dois", "doi_table") |>
  arrow::open_dataset() |>
  filter(chapter == "Geospatial investigation") |>
  collect()

# extract dois
doi_banksia <- doi_table |> filter(name == "banksia") |> pull(doi)
doi_quokkas <- doi_table |> filter(name == "quokkas") |> pull(doi)

# download
banksia <- galah_call() |>
  filter(doi == doi_banksia) |>
  atlas_occurrences()

quokkas <- galah_call() |>
  filter(doi == doi_quokkas) |>
  atlas_occurrences()
```

```{r prereq-external}
#| echo: false
test <- glue::glue(
"
# packages
library(galah)
library(dplyr)
galah_config(email = \"your-email-here\") # ALA-registered email

banksia <- galah_call() |>
  filter(doi == \"{doi_banksia}\") |>
  atlas_occurrences()

quokkas <- galah_call() |>
  filter(doi == \"{doi_quokkas}\") |>
  atlas_occurrences()
"
)
```

```{r prereq-printed}
#| eval: false
#| code: !expr test
#| echo: true
```

::: aside
<img src="https://ala-images.s3.ap-southeast-2.amazonaws.com/store/f/6/1/1/401b8f4d-d896-493c-9c93-e0aecca9116f/original" class="rounded"/></img>

::: figure-caption
[*Banksia serrata*.<br>Photo by kate_and_sam CC-BY-NC 4.0
(Int)](https://biocache.ala.org.au/occurrences/7a57075b-57fa-4b56-9565-29e8aea08223)
:::

<img src="https://ala-images.s3.ap-southeast-2.amazonaws.com/store/9/b/a/d/d994e948-cdb1-4e33-a925-dfce99bbdab9/original" class="rounded"/></img>

::: figure-caption
[*Setonix brachyurus* munching on leaves.<br>Photo by Stephen Cox
CC-BY-NC 4.0
(Int)](https://biocache.ala.org.au/occurrences/077bb0ae-3ecc-451a-a216-2bf0ebed7954)
:::
:::

<!-- Queries -->
:::{.callout-note collapse="true" appearance="minimal"}

#### Original download queries

**Note:** You don't need to run this code block to read this chapter. It can, however, be useful to see the original download query. This code will download the latest data from the ALA, which you are welcome to use instead, though the data might not exactly reproduce results in this chapter.

```{r prereq-orig-query}
#| eval: false
library(galah)
galah_config(email = "your-email-here")

banksia <- galah_call() |>
  identify("banksia serrata") |>
  filter(year > 2022) |>
  select(group = "basic",
         coordinatePrecision, 
         coordinateUncertaintyInMeters) |>
  atlas_occurrences() # <1>

quokkas <- galah_call() |>
  identify("Setonix brachyurus") |>
  galah_select(group = "basic", 
               dataGeneralizations) |>
  atlas_occurrences() # <1>
```
1. We created a custom DOI for our download by using `atlas_occurrences(mint_doi = TRUE)`.

:::

::: {.callout-important collapse="true"}
## Using "Assertions"

<!-- I think it might be helpful here to clarify the wording to be explicit about whether an assertion is the actual data quality check, or a column in the database that flags DQ after running checks, or both - SB -->

To check data quality, data infrastructures like the Atlas of Living
Australia have *assertions*---columns that data infrastructures use to
flag when a record has an issue that fails a data quality check. If we
use galah to download records, we can use assertions columns in our
query to help identify and clean suspicious records.

If you would like to view assertions, use `show_all()`.

```{r}
assertions <- show_all(assertions)
assertions |>
  print(n = 7)
```

You can use the stringr package to search for text matches.

```{r}
assertions |>
  filter(str_detect(id, "COORDINATE")) |>
  print(n = 5)
```

In this chapter, we will detail when an assertion column can help
identify occurrence records with geospatial issues.
:::

## Quick visualisation

Mentioned in the [Inspect chapter](../1_exploring/inspect.qmd), one of the most
straightforward ways to check for spatial errors is to plot your data
onto a map. More obvious spatial errors are much easier to spot
visually.

In most spatial datasets, the most important columns are
`decimalLatitude` and `decimalLongitude` (or similarly named columns).
These contain the latitude and longitude of each observation in decimal
form (rather than degrees).

```{r}
# Retrieve map of Australia
aus <- st_transform(ozmap_country, 4326)

# A quick plot of banksia occurrences
ggplot() + 
  geom_sf(data = aus, colour = "black", fill = NA) + 
  geom_point(data = banksia, 
             aes(x = decimalLongitude,
                 y = decimalLatitude),
             colour = "orchid")
```

<!-- I think it would be helpful to expand on this slightly: we've told readers to plot observations on a map to look for erros, but it may not be intuitive to know what eactly to look for in the first instance. I don't know if there are any errors here worth pointing out, but if not, maybe just a couple of lines to explain what we quickly eyeballed to check that it looks right e.g. no records in the ocean, restricted to the SE coast (are Tas records valid?) - SB -->

## Precision

Not all observations have the same degree of precision. Coordinate
precision can vary between data sources and recording equipment. For
example, coordinates recorded with a GPS unit or a phone generally have
higher precision than coordinates recorded manually from a locality
description.

The degree of precision you require will depend on the granularity of
your research question and analysis. A fine-scale question will require
data measured at a fine-scale to answer it. National or global scale
questions require less precise data.

When downloading data from the ALA with the galah package, it's possible to
include the
[`coordinatePrecision`](https://dwc.tdwg.org/terms/#dwc:coordinatePrecision)
field with your data; this provides a decimal representation of the precision of
coordinates for each observation.

```{r}
banksia |>
  select(scientificName, 
         coordinatePrecision
         ) |>
  filter(!is.na(coordinatePrecision)) # <1>
```

1.  Not all records have this information recorded, so we also filter to
    only records with a `coordinatePrecision` value.

Only a few records have `coordinatePrecision` recorded, but that subset
of records are very precise.

```{r}
banksia |> 
  group_by(coordinatePrecision) |>
  count()
```

<!-- Could potentially remove this step; I don't think it adds much - SB -->

Filter your records to only those under a specific measure of precision.

```{r}
#| eval: false
# Filter by number of decimal places
banksia <- banksia |>
  filter(coordinatePrecision <= 0.001)
```

## Uncertainty

<!-- I got a little confused here initially, and I wonder if it would be less (potentially) confusing to refer to precision and accuracy instead, and explain that accuracy is referred to as uncertainty in the ALA? I'm still not even sure I have correctly understood this, so feel free to ignore if I've gotten it wrong - SB -->

Similarly, not all observations have the same degree of location
certainty. An organism's *exact* location will likely have an area of
uncertainty around it, which can grow or shrink depending on the method
of observation and the species observed. For example, if a person uses a
precise GPS and high-definition camera to make observations of snails,
their observed location will be of higher certainty than a person using
a phone camera to make observations of birds in the distance. The exact
location might also be obscured [for sensitivity
purposes](geospatial-investigation.qmd#obscured-location).
Uncertainty inevitably affects [how robust the results from species
distribution models
are](https://besjournals.onlinelibrary.wiley.com/doi/pdfdirect/10.1111/j.1365-2664.2007.01408.x).

When downloading data from the ALA with the galah package, it's possible to
include the
[`coordinateUncertaintyInMeters`](https://dwc.tdwg.org/terms/#dwc:coordinateUncertaintyInMeters)
field with your data. This refers to the margin of error, represented as a
circular area, around the true location of the recorded observation. We added
this column in our [original galah
query](geospatial-investigation.qmd#prerequisites).

```{r}
banksia |>
  select(scientificName,
         coordinateUncertaintyInMeters
         )
```

There is a range of coordinate uncertainty in our data, with many
falling within 10m of uncertainty.

```{r}
banksia |> 
  count(coordinateUncertaintyInMeters) 
```

<!-- Similarly, not sure this step adds much here - SB -->

If your analysis requires greater certainty, you can then filter your
records to a smaller area of uncertainty.

```{r}
#| eval: false
# Filter by number of decimal places
banksia <- banksia |>
  filter(coordinateUncertaintyInMeters <= 5)
```

## Obscured location

Occurrence records of sensitive, endangered, or critically endangered
species may be obscured (i.e. generalised, obfuscated) to protect the
true location of the species. This process blurs the of an organism
actual location to avoid risks like poaching or capture while still
allowing their data to be included in broader summaries.

In the ALA, the field `dataGeneralizations` contains information of
whether a record has been has been obscured and the size of the area the
point has been generalised to.

::: {.callout-note collapse="true"}
#### Note on `dataGeneralizations`

The `dataGeneralizations` field will only be available to use or
download when there are records in your query that have been
generalised/obscured.
:::

```{r}
search_all(fields, "dataGeneralization")
```


```{r}
#| echo: false

total_tortoise_count <- galah_call() |>
  identify("Pseudemydura umbrina") |>
  atlas_counts()

obscured_tortoise_count <- galah_call() |>
  identify("Pseudemydura umbrina") |>
  group_by(dataGeneralizations) |>
  atlas_counts()
  
```


For example, the Western Swamp Tortoise is [a critically endangered
species in Western
Australia](https://www.dcceew.gov.au/environment/biodiversity/threatened/action-plan/priority-reptiles/western-swamp-tortoise).
There are `r total_tortoise_count` records of this species in the ALA.

```{r}
galah_call() |>
  identify("Pseudemydura umbrina") |>
  atlas_counts()
```

Grouping record counts by the `dataGeneralizations` column shows that 
`r obscured_tortoise_count$count` of the `r total_tortoise_count` records have been obscured by 10 km.

```{r}
#| column: body-outset-right
galah_call() |>
  identify("Pseudemydura umbrina") |>
  group_by(dataGeneralizations) |>
  atlas_counts()
```

#### What do obscured data look like?

Quokka data offer a nice example to get an idea of what to look for when
data points have been obscured. When plotted, obscured occurrence data
appears as if points were plotted onto a grid
[^geospatial-investigation-1].

[^geospatial-investigation-1]: This is pretty much what actually
    happened---locations have been "snapped" onto a grid determined by
    the generalised distance.

```{r}
# remove records with missing coordinates
quokkas <- quokkas |>
  tidyr::drop_na(decimalLatitude, decimalLongitude)

# aus map
aus <- ozmap_country |> st_transform(4326)

# map quokka occurrences
ggplot() + 
  geom_sf(data = aus, colour = "black", fill = NA) + 
  geom_point(data = quokkas, 
             aes(x = decimalLongitude,
                 y = decimalLatitude,
                 colour = dataGeneralizations |>
                   str_wrap(18))) +
  scale_colour_manual(values = c("sienna3", "snow4"),
                      guide = guide_legend(position = "bottom")) +
  guides(colour = guide_legend(title = "Data\ngeneralizations")) +
  xlim(114,120) + 
  ylim(-36,-31)
```

Keep in mind that survey data can also appear gridded if survey
locations were dispersed at equal distance, so be sure to double check
before assuming data has been obscured!

For more information, check out ALA's [support article about working
with threatened, migratory and sensitive
species](https://support.ala.org.au/support/solutions/articles/6000261705-working-with-conservation-and-sensitive-species-information).

## Summary

In this chapter, we showed some ways to investigate the geospatial
coordinates of your data and determine its level of precision,
uncertainty or obscurity.

In the next chapter, we'll see examples of issues with coordinates that
require fixing or removing.
